STAT 672: Homework 3

### 2018-03-21 hw3/hw3.tex
final commit

In the inefficient method, we redo the entire singular value decomposition every
time we update $\lambda$. SVD has a computational complexity on the order of
$O(nd^2)$ (with $n$ corresponding to the number of rows of the feature matrix and 
$d$ corresponding to the number of columns, i.e. the dimensionality of the
data). We then multiply $\mathbf{VD}_\lambda \mathbf{U}^T\mathbf{y}$.
A matrix-matrix product $C=AB$, where $A \in \mathbb{R}^{m \times n}$ and $B \in
\mathbb{R}^{n \times p}$, costs $O(2mnp)$ flops. In our case, $\mathbf{U}^T \in
\mathbb{R}^{d \times n}$ and $\mathbf{Y} \in \mathbb{R}^{n \times 1}$, and so
multiplying the two costs $O(2 \times n \times d \times 1) = O(2nd)$ flops and results
in a $d \times 1$ matrix. Multiplying $\mathbf{D}_\lambda \in \mathbb{R}^{d
\times d}$ and this $d \times 1$ matrix costs $O(2d^2)$ flops (assuming we do not take advantage of the
diagonal structure of $\mathbf{D}_\lambda$) and results in a $d \times 1$ matrix. 
Multiplying $\mathbf{V} \in \mathbb{R}^{d \times d}$ by this $d \times 1$ matrix costs 
$O(2d^2)$ flops. Adding together all these steps, we have $O(nd^2 + 2nd + 2d^2 +
2d^2)$. Dropping all constant coefficients and only considering the
highest-order polynomial, we conclude that the inefficient method costs
$O(nd^2)$ flops.

A more efficient method notes that $\mathbf{D}_\lambda$ is the only part of 
$\mathbf{VD}_\lambda \mathbf{U}^T\mathbf{y}$ that depends on $\lambda$
($\mathbf{U}$ and $\mathbf{V}$ depend solely on the feature matrix $\mathbf{X}$, and
$\mathbf{y}$  depends only on itself) and so we do not not need to recompute the
SVD for every new value of $\lambda$. Assume that we have pre-calculated and
cached $\mathbf{U}^T\mathbf{y}$ and $\mathbf{V}$. Modifying the $d$ non-zero
values of $\mathbf{D}_\lambda$ to reflect our new value of $\lambda$ costs $O(d)$ flops. 
Computing new regression coefficients requires multiplying the new $\mathbf{D}_\lambda$ and
$\mathbf{U}^T\mathbf{y}$, which costs $O(2d^2)$ flops, and then multiplying that
result by $\mathbf{V}$, which also costs $O(2d^2)$ flops. Adding together these
two steps, we have $O(d + 2d^2 + 2d^2)$. Dropping all constant coefficients and
only considering the highest-order polynomial, we conclude that the efficient method
costs $O(d^2)$ flops.

\begin{centering}
\begin{table}[h!]
	\begin{tabular}{|l|l|}
		\hline
		& \textbf{Complexity} \\
		\textbf{Inefficient method} & $O(nd^2)$ \\
		\textbf{Efficient method} & $O(d^2)$ \\
		\hline
	\end{tabular}
\end{table}
\end{centering}

@@

>>>>

In the inefficient method, we recompute the SVD every
time we update $\lambda$. SVD has complexity on the order of
$O(nd^2)$ (with $n$ corresponding to the number of rows of the feature matrix and 
$d$ corresponding to the number of columns). We then multiply $\mathbf{VD}_\lambda \mathbf{U}^T\mathbf{y}$.
A matrix-matrix product $C=AB$, where $A \in \mathbb{R}^{m \times n}$ and $B \in
\mathbb{R}^{n \times p}$, costs $2mnp$ flops. In our case, $\mathbf{U}^T \in
\mathbb{R}^{d \times n}$ and $\mathbf{Y} \in \mathbb{R}^{n \times 1}$, and so
multiplying the two costs $2 \times n \times d \times 1 = 2nd$ flops and results
in a $d \times 1$ vector. Multiplying $\mathbf{D}_\lambda \in \mathbb{R}^{d
\times d}$ and this $d \times 1$ vector costs $d$ flops (assuming we take advantage of the
diagonal structure of $\mathbf{D}_\lambda$) and results in a $d \times 1$ vector. 
Multiplying $\mathbf{V} \in \mathbb{R}^{d \times d}$ by this $d \times 1$ vector costs 
$2d^2$ flops. Adding together all these steps, we have $nd^2 + 2nd + d +
2d^2$ flops. Dropping all constant coefficients and only considering the
highest-order polynomial, we conclude that the inefficient method costs
$O(nd^2)$ flops.

A more efficient method notes that $\mathbf{D}_\lambda$ is the only part of 
$\mathbf{VD}_\lambda \mathbf{U}^T\mathbf{y}$ that depends on $\lambda$
and so we do not need to recompute the
SVD for every new value of $\lambda$. Assume that we have pre-calculated and
cached $\mathbf{U}^T\mathbf{y}$ and $\mathbf{V}$. Modifying the $d$ non-zero
values of $\mathbf{D}_\lambda$ to reflect our new value of $\lambda$ costs
$2d$ flops ($d$ flops for addition of the new $\lambda$ in the denominator, 
and $d$ flops to
divide the numerator by the new denominator). 
Computing new regression coefficients requires multiplying the new $\mathbf{D}_\lambda$ and
$\mathbf{U}^T\mathbf{y}$, which costs $2d^2$ flops, and then multiplying that
result by $\mathbf{V}$, which also costs $2d^2$ flops. Adding together these
two steps, we have $2d + 2d^2 + 2d^2$ flops. Dropping all constant coefficients and
only considering the highest-order polynomial, we conclude that the efficient method
costs $O(d^2)$ flops.

In practical terms, since we are fitting regression coefficients on a training
dataset with $n=6000$, the efficient method is 6,000 times faster.

@@

In the below graph, I have omitted some tested values of $\lambda$ to make
things more visually clear.
